---
author: "Brock"
date: "2025-11-17"
output: rmarkdown::html_vignette
title: Accessing and Modifying Key Columns
vignette: >
  %\VignetteIndexEntry{Accessing and Modifying Key Columns}
  %\VignetteEngine{knitr::knitr}
editor_options:
  chunk_output_type: console
---

# Accessing and Modifying Key Columns

## Setup

Starting with our df_mini, let's rename columns to truly custom names.
```{r}
library(trackframe)

df <- trackframe::df_mini

df

colnames(df) <- c('when', 'up', 'right', 'who')
```

Now let's convert this dataframe to a `trackframe` object, telling `as.trackframe` which columns are the key columns.
```{r}

tf <- as.trackframe(
  df,
  time_col = 'when',
  easting_col = 'right',
  northing_col = 'up',
  id_col = 'who',
  crs = NA
)
tf
```

## Accessing key contents

Without changing our funky column names, we can access the key columns in a generic way:

```{r}
easting(tf)
northing(tf)
time(tf)
id(tf)
```

## Modifying key columns

We can also update the contents of columns using these same functions:

```{r}
time(tf) <- time(tf) + as.difftime(1, unit = 'days')
tf$when # sucessfully updated
```


## Accessing key column names

If we need to access the column names, we can do that with the *_col functions.

```{r}
time_col(tf)
easting_col(tf)
northing_col(tf)
id_col(tf)
```

Alternatively, we can use the `tf_colnames` function. Note: this will just return the names of key columns, not any other columns in the trackframe. The vector returned by `tf_colnames` also
itself has names indicating which key column is which.

```{r}
tf$why <- "just because"

colnames(tf)
tf_colnames(tf)
```

## Modifying key column names

`tf_colnames()` and the individual `*_col()` functions can be used to change the
column names as well.

```{r}
id_col(tf) <- "track_id"

tf_colnames(tf) # updated


tf_colnames(tf)[["time"]] <- "timestamp"
tf_colnames(tf) # updated
```

## Changing which are key columns

What if you want to change which column within the data frame is identified as a key column.
That can be done by rerunning `as.trackframe` with a new column name. For example:

```{r}
# create new columns that we want to be the new coordinates
tf$up_10x <- 10 * tf$up
tf$right_10x <- 10 * tf$right

# up/right and up_10x/right_10x are both part of the trackframe
colnames(tf)

# still same as original
min(easting(tf))
min(northing(tf))

tf_colnames(tf) # also unchanged

# update which columns are used as key columns

tf <- as.trackframe(tf, easting = "right_10x", northing = "up_10x", crs = NA)

# Now the easting and northing values are coming from the *_10x columns
min(easting(tf))
min(northing(tf))

# likewise, tf_colnames is updated
tf_colnames(tf)

# up and right (1x) columns are still there.
# They are just not identified as the easting and northing columns
colnames(tf)
```

## Accessing and updating crs

`crs()` can be used to access and update crs. `crs_type()` can 

```{r}
crs(tf)
crs_type(tf)
```
```{r}
crs(tf) <- "EPSG:32632"
crs(tf)
crs_type(tf)
```
