---
author: "TODO"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: More stuff that should go somewhere
vignette: >
  %\VignetteIndexEntry{Examples of Trackframe}
  %\VignetteEngine{knitr::knitr}
editor_options:
  chunk_output_type: console
---

# Detailed Considerations

This vignette shows some of the more detailed considerations, and under-the-hood processes, that are important to understand when using **trackframe**. It also shows how to change some of the built-in defaults.

## Load Package

```{r}
library(trackframe)
```

## When creating a `trackframe`

### Column guessing

When creating a `trackframe` object from another data type, if the argument `colnames` is not specified, the columns `time_col`, `easting_col`, `northing_col` and `id_col` are guessed automatically. If there are multiple matches, then the first match is chosen. The guesses for each of the four columns are as follows...


```{r}
tf_options("time_col")
tf_options("easting_col")
tf_options("northing_col")
tf_options("id_col")
```

If needed, these guesses can be modified by supplying calues in `tf_options` can be modified by:

```{r}
tf_options("time_col", c(tf_options("time_col"), "tindex2"))
tf_options("time_col")
```



### Coordinate transformation

FIXME: Update once we have updated tra

All the S3 methods of `as.trackframe()` automatically transform coordinates to UTM eastings/northings. If not stored in the input object, information `crs_input` needs to be provided (if `crs_input = NULL` , then `4326` is used as the default crs). An output UTM EPSG can be specified, using the `utm_epsg` argument. If `utm_epsg` is not specified, then it is calculated from the latitude/longitude coordinates by default.

Here is an example using a move2 object:

```{r}
library(move2)

# Load a move2 object using the move2 package
albatross_move2 <- mt_read(mt_example())
head(albatross_move2)
# Original coordinate system is WGS 84 (i.e., crs = 4326)

# Convert to trackframe
albatross_tf <- as.trackframe(data = albatross_move2)

# Extract the input crs EPSG code from the tf object
attributes(albatross_tf)$transformation_info$crs_code

# Extract the projected UTM EPSG code from the tf object
attributes(albatross_tf)$utm_epsg

```

Here is an example using a data.frame object:

```{r}

# Create some example data

df <- data.frame(
  time_col = rep(as.POSIXct(Sys.time() + 1:5), times = 5),
  easting_col = runif(25, 0, 10),
  northing_col = runif(25, 0, 10),
  id = rep(1:5, each = 5)
)

# Convert to trackframe
tf <- as.trackframe(df)

# Extract the input crs EPSG code from the tf object
attributes(tf)$transformation_info$crs_code

# Extract the projected UTM EPSG code from the tf object
attributes(tf)$utm_epsg
```
