---
author: "TODO"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: Detailed Considerations
vignette: >
  %\VignetteIndexEntry{Detailed Considerations}
  %\VignetteEngine{knitr::knitr}
editor_options:
  chunk_output_type: console
---

# Detailed Considerations

This vignette shows some of the more detailed considerations, and under-the-hood processes, that are important to understand when using **trackframe**. It also shows how to change some of the built-in defaults.

## Load Package

```{r}
library(trackframe)
```

## When creating a `trackframe`

### Column guessing

When creating a `trackframe` object from another data type, if the argument `colnames` is not specified, the columns `time_col`, `easting_col`, `northing_col` and `id_col` are guessed automatically. If there are multiple matches, then the first match is chosen. The guesses for each of the four columns are as follows...


```{r}
tf_options("time_col")
tf_options("easting_col")
tf_options("northing_col")
tf_options("id_col")
```

If needed, these guesses can be modified by supplying calues in `tf_options` can be modified by:

```{r}
tf_options("time_col", c(tf_options("time_col"), "tindex2"))
tf_options("time_col")
```



### Compatible Coordinate Systems

`trackframe` only supports cartesian/eucliean coordinate systems ie coordinate systems
with the same units on both axises. For georeferenced coordinates, this means projected coordinates.
For sf objects (move2 and sftrack), `trackframe` identifies the coordinate system using `sf::st_crs`.
For non-sf objects, the user must specify the coordinate system with the `crs` arg in `as.trackframe`.
`NA` can be used to indicate non-georeferenced coordinate systems which still must be cartesian.
(Although we have no way to enforce this.)

For this reason, `trackframe` may not be appropriate for studying processes that cover large
enough portions of the earth that distortion from projected coordinates causes meaningful inaccuracy.


```{r, eval = TRUE}
library(move2)

# Load a move2 object using the move2 package
albatross_move2 <- mt_read(mt_example())

# Original coordinate system is WGS 84 (i.e., crs = 4326)
sf::st_crs(albatross_move2)
```

```{r, eval = TRUE, error = TRUE}
# Convert to trackframe
albatross_tf <- as.trackframe(data = albatross_move2)
```

Fails as this is lat lon.

```{r, eval=TRUE}
albatross_tf <- albatross_move2 |>
  sf::st_transform(32738) |>
  as.trackframe()

# Extract the input crs EPSG code from the tf object
attributes(albatross_tf)$transformation_info$crs_code

```

Here is an example using a data.frame object:

```{r eval = TRUE}

# Create some example data

df <- data.frame(
  time_col = rep(as.POSIXct(Sys.time() + 1:5), times = 5),
  easting_col = runif(25, 0, 10),
  northing_col = runif(25, 0, 10),
  id = rep(1:5, each = 5)
)

# Convert to trackframe with unspecified crs
tf <- as.trackframe(df, crs = NA)

# Trackframe with specified crs
tf <- as.trackframe(df, crs = "EPSG:32632")

# Extract the input crs EPSG code from the tf object
attributes(tf)$transformation_info$crs_code
```
