---
author: "Alison Ashbury & Brock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: Why trackframe
vignette: >
  %\VignetteIndexEntry{Why trackframe}
  %\VignetteEngine{knitr::knitr}
editor_options:
  chunk_output_type: console
---

# Why trackframe?

The trackframe package is designed to be complementary to more full featured geospatial libraries such as [sf](https://r-spatial.github.io/sf/). This vignette demonstrates the main advantages of using trackframe rather than non-trackframe approaches to writing functions for manipulating and analyzing animal tracks.

## Main advantages

* The `trackframe` object is a flexible `data.frame` for storing animal track data; coordinates, times, and animal ids, are stored as columns.
* Multiple input object structures (e.g., `data.table`s, `tibble`s, `move2` objects, `sftrack` objects) can be easily coerced into `trackframe` objects.
* A `trackframe` object can be easily backtransformed into its original data structure using `tf_backtransform()`.
* Key columns, namely the x (easting) and y (northing) coordinates of each point, as well as the timestamp (date & time) and unique animal id, are stored as attributes and can be easily accessed by `easting()`, `northing()`, `time()`, and `id()`, regardless of the columns' actual names.
* Coordinates are always stored as projected UTM easting/northing (for georeferenced data) or as x/y values on a Cartesian plane (for non georeferenced data, e.g., simulation data).
* Coordinates are stored in simple numeric columns, allowing the user to perform vectorized R operations directly on the coordinates themselves
* Fast performance.
* Trackframe's objects and functions integrate seamlessly with the travelpaths package.

## Example

Below is example function that takes x, y, t data columns from a single animal track as inputs, and outputs the average travel speed of the animal. We want to write code that applies this function to multiple tracks (for e.g., separate tracks for different individual animals) provided in a single object.

```{r}
average_speed_over_time <- function(x, y, t)  {
  weighted.mean(sqrt(diff(x)^2 + diff(y)^2) / as.numeric(diff(t)),
    as.numeric(diff(t)), na.rm = TRUE)
}
```

### Without trackframe

Here is an example of how this would be done without trackframe, if you wanted the function to be able to take a variety of different data representations as input objects (e.g., `move2` object, `sftrack` object, `data.frame`, etc).

```{r}
average_speed_group_no_tf <- function(data) {
  # Both move2 and sftrack are sf based
  if ("sf" %in% class(data)) {
    # sf objects have a consistent way of providing coordinates
    # we need to project them into euclidean space

    # coords <- sf::st_coordinates(sf::st_transform(data, 3857))
    stopifnot(!sf::st_is_longlat(sf::st_crs(data)))
    coords <- sf::st_coordinates(data)
    coords[is.nan(coords)] <- NA
    # sftrack and move2 have different ways of encoding
    # the track_id/animal_id so we need to account for both
    if ("sftrack" %in% class(data)) {
      track_id <- sapply(data[[attr(data, "group_col")]], \(x) x[["id"]])
      timestamp <- data[[attr(data, "time_col")]]
    } else if ("move2" %in% class(data)) {
      track_id <- move2::mt_track_id(data)
      timestamp <- move2::mt_time(data)
    }
    # This part is for the vanilla dataframe
  } else {
    coords <- as.matrix(data[, c("x", "y")])
    track_id <- data$animal_id
    timestamp <- data$t
  }
  #  At this point, we have standardized coords, timestamp, and track_id objects

  # Now we can finally use sapply to run our `average_speed_over_time`
  # function for each track.
  utid <- sort(unique(track_id))
  names(utid) <- utid
  sapply(
    utid,
    function(focal_track_id) {
      idx <- which(track_id == focal_track_id)
      average_speed_over_time(coords[idx, 1], coords[idx, 2], timestamp[idx])
    }
  )
}
```

### With trackframe

Using the `as.trackframe` function, we can pull in data of many other types, and trackframe will automatically detect which columns are `easting` (i.e., x or longitude), `northing` (i.e., y or latitude), `time` (i.e., the date-time stamp), and `id` (i.e., the identifier of each individual animal). We can then extract those columns as vectors using the `easting()`, `northing()`, `time()`, and `id()` functions respectively. No custom logic dictated by the input format, nor juggling of original column names, is needed to accomplish this.

```{r}
library(trackframe)

average_speed_group_tf <- function(data, ...) {

  # Automatically convert whatever input type is provided.
  tf <- as.trackframe(data, ...)
  # id_col='animal_id' is only used for the dataframe conversion.
  # track_id/animal_id is automatically detected from move2 and sftrack objects

  # Run the function over for each animal
  sapply(split(tf, id(tf)), function(tf) {
    # Use trackframe functions to specify the necessary columns
    average_speed_over_time(easting(tf), northing(tf), time(tf))
  })
}
```

### Check output

In the following code, we check that our newly created function (using trackframe) works with the most common formats of animal track data (`data.frame`, `sftrack`, and `move2`), and that it returns the same results as the messy cross-compatible function that does not use trackframe.

```{r}
# dataframe
set.seed(2025)
df <- data.frame(
  x = rnorm(10),
  y = rnorm(10),
  t = 1:10,
  animal_id = c(rep("a", 5), rep("b", 5))
)
stopifnot(all.equal(
  average_speed_group_tf(df, crs = NA),
  average_speed_group_no_tf(df)
))

# sftrack
sftrack_raccoon <- sftrack::as_sftrack(
  data = sftrack::raccoon,
  coords = c("longitude", "latitude"),
  time = "timestamp",
  group = "animal_id",
  crs = 4326
)

sftrack_raccoon_proj <- sf::st_transform(
  sftrack_raccoon,
  trackframe::suggest_utm_crs(sftrack_raccoon)
)

stopifnot(all.equal(
  average_speed_group_tf(sftrack_raccoon_proj),
  average_speed_group_no_tf(sftrack_raccoon_proj)
))

# move2
move2_albatross <- move2::mt_read(move2::mt_example())

move2_albatross_proj <- sf::st_transform(
  move2_albatross,
  trackframe::suggest_utm_crs(move2_albatross)
)

stopifnot(all.equal(
  average_speed_group_tf(move2_albatross_proj),
  average_speed_group_no_tf(move2_albatross_proj)
))
```

## Why non-`sf` based?

Trackframe is designed to be complementary to more full featured geospatial libraries such as sf. The sf library uses a geometry list column to store Euclidean or Haversine coordinates. This allows sf to support many geometry types (`LINESTRING`, `POLYGON` etc.) and facilitates interfacing with non-R geospatial/geometry programs. However, it does not allow the user to perform vectorized R operations directly on the coordinates. In contrast, trackframe is focused on supporting analyses that operate on the geometry and timing of the x, y, t data itself, rather than how the track is embedded in other geospatial systems. To facilitate efficient R code for these use cases, `trackframe` objects store Euclidean coordinates directly as dataframe columns.
