---
author: "Alison Ashbury"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: Compatible Coordinate Systems
vignette: >
  %\VignetteIndexEntry{Compatible Coordinate Systems}
  %\VignetteEngine{knitr::knitr}
editor_options:
  chunk_output_type: console
---

# Compatible Coordinate Systems

Trackframe only supports Cartesian (or Euclidean) coordinate systems, i.e., systems in which coordinates describe locations on a flat plane using perpendicular northing (y) and easting (x) axes, and units represent consistent physical distances in all directions.

As long as the coordinates are Cartesian, trackframe supports both: georeferenced data, wherein the coordinates explicitly represent locations on the Earth's surface; and non-georeferenced data, wherein the coordinates do not explicitly or necessarily represent locations on the Earth's surface.

## Load Package

```{r}
library(trackframe)
```

## Georeferenced data

For georeferenced coordinates (i.e., coordinates that represent locations on the Earth's surface), `trackframe` requires that they be projected, e.g., UTMs rather than latitute-longitude.

For `sf` objects (`move2` and `sftrack` objects), `trackframe` identifies the coordinate system using `sf::st_crs`. For non-sf objects, you have to specify the coordinate reference system with the `crs` argument in `as.trackframe`. If your data has non-Cartesian georeferenced coordinates, for example, latitude and longitude, you will need to first project your data before converting it to to a trackframe object. For this reason, trackframe may not be appropriate for studying movement processes that cross large enough distances that distortion from projected coordinates causes meaningful measurement inaccuracies (e.g., long-distance migration).

Take this example, using a `move2` object:

```{r, eval = TRUE}
library(move2)

# Load a move2 object using the move2 package
albatross_move2 <- mt_read(mt_example())

# Check the coordinate reference system
sf::st_crs(albatross_move2)
```

The objects coordinate reference system is WGS 84 (i.e., `crs = 4326` or `EPSG:4326`), i.e., the standard decimal latitude-longitude. These are, thus, not Cartesian coordinates. If we attempt to directly convert this object to a `trackframe` object, it will fail...

```{r, eval = TRUE, error = TRUE}
# Convert to trackframe (ERROR since non-Cartesian coordinates)
albatross_tf <- as.trackframe(data = albatross_move2)
```

Note the instruction in the error message: "Please project into an appropriate crs."

There are several packages and approaches for projecting georeferenced data. In the following example, we use the `st_transform` function from the `sf` package. 

```{r, eval=TRUE}
# Project the coordinates into the appropriate UTM zone (in this case,
# UTM zone 38S, or EPSG: 32738), and then convert to trackframe object
albatross_tf <- albatross_move2 |>
  sf::st_transform(crs = 32738) |>
  as.trackframe()

# Extract the input crs EPSG code from the tf object's transformation_info
crs(albatross_tf)
```

<!--

[ONCE THIS IS IMPLEMENTED, remove code block above, replace with text and code block below.]

If you don't already know the EPSG code for the UTM zone that your data are in, you can use trackframe's `suggest_utm_zone_crs` function to find out. (Alternatively, or additionally, you can find or verify the UTM zone at <https://epsg.io/>).

```{r, eval=FALSE, include = FALSE}
# nolint start
# Get EPSG code for likely UTM zone
# suggest_utm_zone_crs(albatross_move2)

# Project the coordinates into the appropriate UTM zone (in this case,
# UTM zone 38S, or EPSG: 32738), and then convert to trackframe object
albatross_tf <- albatross_move2 |>
  sf::st_transform(crs = 32738) |>
  as.trackframe()

# Extract the input crs EPSG code from the tf object's transformation_info
crs(albatross_tf)
# nolint end
```
-->

In the following example, we convert a `data.frame` object to a `trackframe` object. The dataframe doesn't contain any georeferencing metadata (i.e., a crs), but let's say that we know that these data were collected in southern Germany and that the coordinates are already projected (i.e., they're in UTM zone 32N, or ESPG:32632). In this case, you can simply tell `as.trackframe` what the crs is, using the `crs` argument.

```{r eval = TRUE}

# Create some example data
df <- data.frame(
  time_col = rep(as.POSIXct(Sys.time() + 1:5), times = 5),
  easting_col = runif(25, 0, 10),
  northing_col = runif(25, 0, 10),
  id = rep(1:5, each = 5)
)

# Specify the crs when converting to trackframe
tf <- as.trackframe(df, crs = "EPSG:32632")

# Extract the input crs EPSG code from the tf object
crs(tf)
```

## Non-georeferenced data

For non-georeferenced coordinates, i.e., those that do not represent locations on the Earth's surface, the `crs` argument in `as.trackframe` should be set to `NA`. This may apply to, for example, data pertaining to simulations (i.e., representing locations on a virtual plane) or captive animal locations (i.e., representing locations in an enclosure, measured relative to some reference point in or near the enclosure). Note that even non-georeferenced coordinates should still be Cartesian, but `trackframe` has no way of verifying this or alerting if this is not the case.

Where `as.trackframe` does not inherit metadata specifying the `crs` from the input object (for example, when the input object is a dataframe), the `crs` argument **must** be provided. Otherwise `as.trackframe` will throw an error.

```{r eval = TRUE, error = TRUE}
# Create some example data
df <- data.frame(
  time_col = rep(as.POSIXct(Sys.time() + 1:5), times = 5),
  easting_col = runif(25, 0, 10),
  northing_col = runif(25, 0, 10),
  id = rep(1:5, each = 5)
)

# Convert df to trackframe without specifying crs (ERROR since no crs given)
tf_non_gr <- as.trackframe(df)

# Convert df to trackframe with non-georeferenced crs
tf_non_gr <- as.trackframe(df, crs = NA)

# Extract the crs from the tf object
crs(tf_non_gr)
```
