---
author: "Alison Ashbury"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: Converting between trackframe and other data representations
vignette: >
  %\VignetteIndexEntry{Converting between trackframe and other data representations}
  %\VignetteEngine{knitr::knitr}
editor_options:
  chunk_output_type: console
---

# Converting between `trackframe` and other data representations

Here, we show how to convert between some less common animal track data representations and `trackframe` objects using the `as.trackframe` function. For more common animal track data representations (including plain (base R) `data.frame`, `move2`, and `sftrack` objects), check out the [Quickstart vignette](trackframe.html).

<!-- Right now, it's just cocomo, but i figure maybe we want to add here in the future? or should this just be cocomo? if so, easy to change, just let me know! -->

## Converting to and from a `cocomo` dataset 

The [`cocomo` R package](https://livingingroups.github.io/cocomo/) was developed in the [CoCoMo research group](https://www.ab.mpg.de/cocomo) at the Max Planck Institute of Animal Behavior. The package has many functions for the analysis of communication and movement of animal groups. 

`cocomo` functions require that animal track data be stored in 4 separate objects: 2 `matrix` objects, one each representing the x and the y positions of all individuals, 1 `vector` of timestamps, and 1 `dataframe` containing ids and information about the individuals. We will refer to this dataset structure as a "cocomo dataset" throughout this vignette.

### Convert a `trackframe` object to a cocomo dataset

To transform a `trackframe` object into a cocomo dataset, use the `tf_as_cocomo` function. This function returns a single `list` that contains the 4 objects that make up a cocomo dataset of animal tracks.

```{r}
library(trackframe)

# Create a trackframe object
tf <- as.trackframe(data.frame(
  timestamp = rep(as.POSIXct(Sys.time() + 1:7), times = 3),
  x = runif(21, 0, 10),
  y = runif(21, 0, 10),
  id = rep(1:3, each = 7)),
  crs = NA)

# Convert tf to a cocomo dataset
cocomo_list <- tf_as_cocomo(tf)

# Check out the cocomo dataset
attributes(cocomo_list)
str(cocomo_list)
cocomo_list$xs
cocomo_list$ys
cocomo_list$t
cocomo_list$ids
```

So, this cocomo dataset is a `list` that contains 4 objects: the first is a `matrix` of the x (easting) positions of every individual (rows) across all timestamps (columns); the second is a `matrix` of the y (northing) positions of every individual across all timestamps, the third is a `vector` of the 7 timestamps, and the fourth is a `dataframe` of the 3 individuals' ids. 

We can then use a function from the `cocomo` package on the objects in this cocomo dataset, for example:

```{r}
# Install cocomo package
library(devtools)
install_github('livingingroups/cocomo')
library(cocomo)

# Get the group centroid at each time step
get_group_centroid(xs = cocomo_list$xs, 
                   ys = cocomo_list$ys)
```

### Convert a cocomo dataset to a `trackframe` object

To convert a `cocomo` dataset to a `trackframe` object, we can use the dedicated `cocomo_as_tf` function, and specify exactly where to find the information for each of the key columns in the arguments.

```{r}
# Convert cocomo dataset to trackframe object
tf2 <- cocomo_as_tf(
  xs = cocomo_list$xs,
  ys = cocomo_list$ys,
  t = cocomo_list$t,
  ids = cocomo_list$ids,
  utm_epsg = 32601) # <-------------------- Needs update to 'crs = ' with NA option

tf2
class(tf2)
attributes(tf2)[1:6]
```

<!-- FIX ME: Update this after Issue #163 is resolved. -->


