---
author: "Alison Ashbury"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: Converting between trackframe and other data representations
vignette: >
  %\VignetteIndexEntry{Converting between trackframe and other data representations}
  %\VignetteEngine{knitr::knitr}
editor_options:
  chunk_output_type: console
---

# Converting between `trackframe` and other data representations

Here, we show how to convert between some less common animal track data representations and `trackframe` objects using the `as.trackframe` function. For more common animal track data representations (including plain (base R) `data.frame`, `move2`, and `sftrack` objects), check out the [Quickstart vignette](trackframe.html).

<!-- Right now, it's just cocomo, but i figure maybe we want to add here in the future? or should this just be cocomo? if so, easy to change, just let me know! -->

## Converting to and from `cocomo` objects

The [`cocomo` R package](https://livingingroups.github.io/cocomo/) was developed in the [CoCoMo research group](https://www.ab.mpg.de/cocomo) at the Max Planck Institute of Animal Behavior. The package has many functions for the analysis of communication and movement of animal groups. These functions require that the data be in a special `list` of `matrix` objects. 

<!-- Not sure what to use as a starting point here - cocomo package doesn't seem to have any example data, so i'm gonna make some via tf, but hide it. alternatively, if we can get some og cocomo data for this, or if that exists in the package and i missed it, that would be better!-->

```{r, eval = TRUE, include = FALSE}
library(cocomo)

# in the background, create a cocomo matrix obj to use as an example
tf_cocomo <- as.trackframe(data.frame(
  timestamp = rep(as.POSIXct(Sys.time() + 1:7), times = 3),
  x = runif(21, 0, 10),
  y = runif(21, 0, 10),
  id = rep(1:3, each = 7)
))
cocomo_obj <- tf_as_cocomo(tf_cocomo)

```

The following toy `cocomo` object has the x and y positions of 3 individuals over 7 sequential timestamps. The x and y positions are stored in matrices called `xs` and `ys`, with one row per individual and one column per timestamp. The timestamps are stored in a vector called `t`, and the ids are stored in a small `data.frame` called `ids`.

```{r}
library(cocomo)

# Check out the cocomo object
attributes(cocomo_obj)
str(cocomo_obj)
cocomo_obj$xs
cocomo_obj$ys
cocomo_obj$t
cocomo_obj$ids
```

### Create a `trackframe` from a `cocomo` object

To convert a `cocomo` object to a `trackframe` object, we can use the dedicated `cocomo_as_tf` function, and specify where to find the information for each of the key columns.

```{r}
# Convert cocomo object to trackframe object
cocomo_tf <- cocomo_as_tf(cocomo_obj,
                           xs = cocomo_obj$xs,
                           ys = cocomo_obj$ys,
                           t = cocomo_obj$t,
                           ids = cocomo_obj$ids,
                           crs = 32632)


cocomo_tf
class(cocomo_tf)
attributes(cocomo_tf)[1:6]
```


### Create a `cocomo` object from a `trackframe`

To transform a `trackframe` object into a `cocomo` object, use the `tf_as_cocomo` function. 

```{r}
ap1_recovered_cocomo <- tf_as_cocomo(cocomo_tf)
head(ap1_recovered_cocomo)
attributes(ap1_recovered_cocomo)

# Compare to original cocomo object
head(cocomo_obj)
attributes(cocomo_obj)
```


