---
author: "TODO"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example of Trackframe vs move2 performance}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  chunk_output_type: console
---



## Why `trackframe`?
TODO

- The R package `trackframe` is designed to be complementary to more full featured geospatial libraries such as `sf`.
- `trackframe` offers a flexible object for tracking analysis with an underlying `data.frame`, `data.table`, or `tibble` object where the information about coordinates, time, id , etc. are stored as columns.
- multiple input objects can be easily coerced to a `trackframe`
- information like `time`, `id`, `easting` and `northing` are stored as attributes an can be easily accessed by `time()`, `id()`, `easting()`, `northing`.
- coordinates are always stored in UTM easting/northings
- a `trackframe` can be easily be backtransformed to the original data by `tf_backtransform()`
- easy transformation allow to use methods from multiple packages which require different input formats. Several methods are implemented in package `travelpaths`.
- can easily deal with multiple individuals
- performance
- allow the user to perform vectorized R operations on directly on the coordinates



## Performance
TODO

### Example of Trackframe vs move2 performance

```{r}
library(trackframe)
library(sf)
```


Distance function for trackframe.

```{r}
library(units)
tf_distance <- function(tf) {
  x <- c(sqrt(diff(easting(tf))^ 2 + diff(northing(tf))^2), NA)
  x[diff(as.numeric(id(tf))) != 0] <- NA
  set_units(x,'m')
}
filter_nonempty <- function(x) x[!sf::st_is_empty(x),]

brock_mt_distance <- function(m2) {
  coords <- sf::st_coordinates(m2)
  x <- sqrt(diff(coords[,1])^2 + diff(coords[,2])^2)
  x[diff(as.numeric(mt_track_id(albatross_move2))) != 0] <- NA
  set_units(c(x, NA), 'm')
}
```

Load data

```{r, results="hide"}
library(move2)
suppressPackageStartupMessages(library(move2, quietly = TRUE))
albatross_move2 <- mt_read(mt_example()) |>
  sf::st_transform(32618) |>
  filter_nonempty()

# utm_epsg <- trackframe:::sf_to_utm_epsg(albatross_move2)
# albatross_move2 <- sf::st_transform(albatross_move2, utm_epsg)

suppressPackageStartupMessages(library(trackframe, quietly = TRUE))
albatross_tf <- as.trackframe(albatross_move2)

all.equal(st_coordinates(albatross_move2)[,1], easting(albatross_tf))
```

```{r}
all.equal(brock_mt_distance(albatross_move2), mt_distance(albatross_move2), check.attributes = FALSE)
```

Check that the distance functions output the same result:
```{r}
all.equal(mt_distance(albatross_move2), tf_distance(albatross_tf), check.attributes = FALSE)
```

Compare performance

```{r}
library(microbenchmark)
m <- microbenchmark(
  mt_distance(albatross_move2),
  brock_mt_distance(albatross_move2),
  tf_distance(albatross_tf),
  times = 15
)
m
plot(m)
```



