---
title: Plotting
author: "Brock"
date: "2025-12-01"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Plotting

## Default Plot Method

Running the `plot` function on a `trackframe` creates a plot of the location
points, joined by lines in temporal order, with easting on the x-axis and northing on the y-axis. 
By default, all tracks are plotted with an aspect ratio of 1, each in a different color. 
The start point of each track is marked by an arrowhead,
and the end point of each track is marked by a bar. 

```{r, include=FALSE}
# supposedly 7in is the defautl, but for some reason on macos, defaults to 3in
knitr::opts_chunk$set(fig.width = 7)
knitr::opts_chunk$set(fig.height = 7)
```

```{r}
library(trackframe)
df_demo
tf_demo <- as.trackframe(df_demo, crs = NA)
```

```{r}
plot(tf_demo)
```


## Customizing Default Plot Method

`trackframe`'s plotting functions are based on the `tinyplot` package
(`vignette("introduction", package = "tinyplot")`). This means that it
responds to largely the same parameters as base R plots, and that
`tinyplot`'s syntax can be used.

### Controling Visible Elements

Each element of the plot (lines, points, start indicators and end
indicators) can be turned on and off.

For example, if you want to show lines only:

```{r}
plot(tf_demo, points = FALSE, start_indicator = FALSE, end_indicator = FALSE)
```

Or points only:

```{r}
plot(tf_demo, lines = FALSE, start_indicator = FALSE, end_indicator = FALSE)
```

### Changing the Theme

You can also change the theme of the plot to any of `tinyplot`'s built-in and custom themes. For example:

```{r}
tinyplot::tinytheme("dark")
plot(tf_demo)
```

```{r}
tinyplot::tinytheme("tufte")
plot(tf_demo)
```

```{r}
# Reset theme
tinyplot::tinytheme()
```

The full range of `tinyplot` themes is shown here: (`vignette("themes", package = "tinyplot")`)

### Adjusting Styles

The style of individual elements can also be adjusted by passing parameters
as a list to that element's corresponding `_style` argument.

For example, if you want the points to be large, you can adjust the
`cex` in a list of the `points_style` argument:

```{r}
plot(tf_demo, points_style = list(cex = 3), start_indicator = FALSE, end_indicator = FALSE)
```

If you want to change the line thickness, you can change the 
`lwd` in a list of the `lines_style` argument:

```{r}
plot(tf_demo, lines_style = list(lwd = 8), start_indicator = FALSE, end_indicator = FALSE)
```

These styling lists (and the plot() method itself) take parameters
passable to tinyplot::tinyplot. This includes parameters defined in
`graphics::par` ...

```{r}
plot(tf_demo, col = "blue")
```

... as well as parameters defined directly in `tinyplot::tinyplot`.

Parameters that apply to the entire plot, not just one element, can be
passed directly to `plot(...)` rather than in a `_style` list.

```{r}
plot(tf_demo, legend = FALSE)
```

### Adjusting Indicators

The start and end indicators are arrows. Their style can be adjusted as
well using parameters that apply to arrows. 

For example, you can adjust
the size of the arrowhead with the `length` parameter.

We can make the end indicators big and start indicators
tiny:

```{r}
plot(tf_demo, start_indicator_style = list(length = .05), end_indicator_style = list(length = 0.3))
```

You can also use the `arrowhead_loc` parameter to move the arrowhead
away from the exact start/endpoint of each track toward the
second/second-to-last point.

For example, this puts the start arrowhead 60% of the way between the
first point of each track and the second point of each track, and puts
the end arrowhead, 20% of the way between the last point in each track
and the second-to-last point:

```{r}
plot(
  tf_demo,
  start_indicator_style = list(arrowhead_loc = .6),
  end_indicator_style = list(arrowhead_loc = .2)
)
```

## Plotting Markers

If you want to highlight certain non-endpoints, create an indicator
column with the points you'd like to highlight marked as `TRUE` and all
others marked as `FALSE`. Then, pass the name of the column to the
`plot()` method as the `marker` arg.

```{r}
# Create an indicator column, all points marked as FALSE
tf_demo$track_1_2nd_point <- FALSE

# Mark the 2nd element of track_1 as TRUE
tf_demo$track_1_2nd_point[id(tf_demo) == "track_1"][2] <- TRUE

plot(tf_demo, marker = "track_1_2nd_point")
```

By default, makers appear as boxes around the relevant datapoint. Similar to other elements,
they can be styled with `marker_style`.

```{r}
plot(tf_demo, marker = "track_1_2nd_point", marker_style = list(pch = "2", cex = 10))
```
