---
author: "Brock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Basic Functionality}
---

# Qucikstart Vignette

This vignette demonstrates how to use `trackframe` to make your functions compatible with multiple animal movement data formats.


Below is example algorithm that takes x, y, t from a single track as inputs. We want to write code that applies this function to multiple tracks provided in a single object.
```{r}
average_speed_over_time <- function(x, y, t)  {
  weighted.mean(sqrt(diff(x)^2 + diff(y)^2) / as.numeric(diff(t)),
    as.numeric(diff(t)), na.rm=TRUE)
}
```


Here is an example of how this would be done without trackframe.
```{r}
average_speed_group_no_tf <- function(data) {
  # Both move2 and sftrack are sf based.
  if ('sf' %in% class(data)) {
    # sf objects have a consistent way of providing coordinates
    # we need to project them into euclidean space

    # coords <- sf::st_coordinates(sf::st_transform(data, 3857))
    utm_epsg <- trackframe:::sf_to_utm_epsg(data)
    coords <- sf::st_coordinates(sf::st_transform(data, utm_epsg))
    coords[is.nan(coords)] <- NA
    # sftrack and move2 have different ways of encoding
    # the track_id/animal_id so we need to account for both
    if ('sftrack' %in% class(data)) {
      track_id <- sapply(data[[attr(data, "group_col")]], deparse)
      timestamp <- data[[attr(data, 'time_col')]]
    } else if ('move2' %in% class(data)) {
      track_id <- move2::mt_track_id(data)
      timestamp <- move2::mt_time(data)
    }
    # This part is for the vanilla dataframe
  } else {
    coords <- as.matrix(data[,c('x', 'y')])
    track_id <- data$animal_id
    timestamp <- data$t
  }
  #  At this point, we have standardized coords, timestamp, and track_id objects
  
  # Now we can finally use sapply to run our `average_speed_over_time`
  # function for each track.
  utid <- sort(unique(track_id))
  names(utid) <- utid
  sapply(
    utid,
    function(focal_track_id){
      idx <- which(track_id == focal_track_id)
      average_speed_over_time(coords[idx,1], coords[idx,2], timestamp[idx])
    }
  )
}
```

Here is how to accomplish the same thing with `trackframe`. No custom logic is needed for each input type. 

```{r}
library(trackframe)

average_speed_group_tf <- function(data, ...){
  # Automatically convert whatever input type is provided.
  tf <- as.trackframe(data, ...)
  # id_col='animal_id' is only used for the vanilla dataframe conversion.
  # track_id/animal_id is automatically detected from move2 and sftrack objects
  
  # Run the function itself over for each animal.
  sapply(split(tf, id(tf)), function(tf) {
    average_speed_over_time(easting(tf), northing(tf), time(tf))
  })
}
```
In the snippet above, `id()`, `easting()`, `northing()` and `time()` are used to pull out track_id, x, y, and t respectively from the trackframe object. No need to explicitly juggle column names.


Doublecheck that our newly created function works with all three formats and returns the same results as the messy cross-compatible function
```{r}
# Vanilla Dataframe
set.seed(2025)
vanilla_df <- data.frame(
  x = rnorm(10),
  y = rnorm(10),
  t = 1:10,
  animal_id = c(rep('a', 5), rep('b', 5))
)
stopifnot(all.equal(
  average_speed_group_tf(vanilla_df),
  average_speed_group_no_tf(vanilla_df)
))

# sftrack
sftrack_raccoon <- sftrack::as_sftrack(
  data = sftrack::raccoon,
  coords = c("longitude","latitude"),
  time = "timestamp",
  group = "animal_id",
  crs = 4326
)
stopifnot(all.equal(
  average_speed_group_tf(sftrack_raccoon),
  average_speed_group_no_tf(sftrack_raccoon)
))

# move2
move2_albatross <- move2::mt_read(move2::mt_example()) |>
  sf::st_transform(3857)

stopifnot(all.equal(
  average_speed_group_tf(move2_albatross, id_col = "individual-local-identifier"),
  average_speed_group_no_tf(move2_albatross)
))
```