---
author: "TODO"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples of Trackframe}
  %\VignetteEngine{knitr::knitr}
editor_options: 
---

# Load Package

```{r}
library(trackframe)
```

# Create a `trackframe`

## General
The R package `trackframe` is designed to be complementary to more full featured geospatial libraries such as `sf`. Objects of class `trackframe` offer a flexible object for tracking analysis with an underlying `data.frame`, `data.table`, or `tibble` object where the information about coordinates, time, id , etc. are stored as columns.
In a `trackframe` we always assume to have UTM easting/northing as coordinates.
The S3 method `as.trackframe` (currently available for `data.frame`, `sftrack` and `move2`) automatically transforms coordinates to UTM easting/northing. `time_col`, `easting_col`, `northing_col` and `id_col` are stored as attributes.

### col guessing
If no specific colnames are provided columns `time_col`, `easting_col`, `northing_col` and `id_col` are guessed automatically if no a argument is provided. of multiple matches, the first match is chosen. Default arguments are specified by

```{r}
tf_options("time_col")
tf_options("easting_col")
tf_options("northing_col")
tf_options("id_col")
```

Values in `tf_options` can be modified by
```{r}
tf_options("time_col", c(tf_options("time_col"), "tindex2"))
tf_options("time_col")
```

### coordinates transformation

All the S3 methods of `as_trackframe()` automatically transform coordinates to UTM eastings/northings. If not stored in the input object, information `crs_input` needs to be provided (if `crs_input = NULL` `4326` is used as default crs). `utm_epsg` is optional and is calculated from the latitude/longitude coordinates by default.

## from a `data.frame`

```{r}
df <- data.frame(
  time_col = as.POSIXct(Sys.time() + 1:5),
  easting_col = runif(5, 0, 10),
  northing_col = runif(5, 0, 10),
  id = 1:5
)
tf <- as.trackframe(df, time_col = "time_col", easting_col = "easting_col",
                        northing_col = "northing_col", id_col = "id")
head(tf)
attributes(tf)
```

## from a `move2` object

```{r}
library(move2)
albatross_move2 <- mt_read(mt_example()) |>
  sf::st_transform(3857)
albatross_move2 <- albatross_move2[!sf::st_is_empty(albatross_move2),]
albatross_tf <- as.trackframe(data = albatross_move2)
class(albatross_tf)
```

Backtransform `trackframe` to `move2` object

```{r}

recovered_move2 <- tf_as_move2(tf = albatross_tf[!is.na(northing(albatross_tf)),])
str(recovered_move2)
```

## from an `sftrack` object

```{r}
library(sftrack)
library(trackframe)
data("raccoon", package = "sftrack")
raccoon$month <- as.POSIXlt(raccoon$timestamp)$mon + 1
raccoon$time <- as.POSIXct(raccoon$timestamp, tz = "EST")
coords <- c("longitude","latitude")
group <- list(id = raccoon$animal_id, month = as.POSIXlt(raccoon$timestamp)$mon+1)
time <- "time"
error <- "fix"
crs <- 4326
# create a sftrack object
my_sftrack <- as_sftrack(data = raccoon,
                         coords = coords,
                         group = group,
                         time = time,
                         error = error,
                         crs = crs)

sftrack_tf <- as.trackframe(my_sftrack)
class(sftrack_tf)
attributes(sftrack_tf)
```

Backtransform `trackframe` to `sf` object

```{r}
recovered_sf <- tf_backtransform(sftrack_tf[!is.na(northing(sftrack_tf)),])
recovered_sf <- tf_backtransform(sftrack_tf)
str(recovered_sf)
```


## from `cocomo` style data

```{r}
set.seed(2025)
data(tf_mini)
cocomo <- tf_as_cocomo(tf_mini)
str(cocomo)
cocomo$x
cocomo$y
cocomo$t
cocomo$ids

cocomo_tf <- cocomo_as_tf(cocomo$x, cocomo$y, cocomo$t, cocomo$ids)
cocomo_tf
class(cocomo_tf)


```

Retransform `trackframe` to cocomo style data

```{r}
cocomo2 <- tf_as_cocomo(cocomo_tf)
str(cocomo2)
```



# Extract information

```{r}
tf <- tf_mini
```


```{r}
easting(tf)
```

```{r}
northing(tf)
```

```{r}
time(tf)
```

```{r}
id(tf)
```