---
author: "Brock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: Quickstart Guide
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Quickstart Guide}
---

# Quickstart Guide

This vignette demonstrates the main functions of **trackframe** and how to get started using this package. For a full description of the benefits of using **trackframe** over other approaches, see [Why trackframe?](#%7BWhy%20trackframe%7D).

## Load Package

```{r}
library(trackframe)
```

## Create a `trackframe`

The following shows how to convert common animal track data types into `trackframe` objects. To determine which columns are relevant to **trackframe**, there is a column guessing process. For more information, see [advanced vignette / specific considerations].

### ...from a `data.frame`

```{r}
# Create some example data
df <- data.frame(
  time_col = rep(as.POSIXct(Sys.time() + 1:5), times = 5),
  easting_col = runif(25, 0, 10),
  northing_col = runif(25, 0, 10),
  id = rep(1:5, each = 5)
)

# Convert to trackframe with explicit column naming
tf <- as.trackframe(
  df, time_col = "time_col", easting_col = "easting_col",
  northing_col = "northing_col", id_col = "id", crs = "EPSG:32632"
)

# Convert to trackframe without naming columns, relying instead on column guessing
tf <- as.trackframe(df, crs = "EPSG:32632")

head(tf)
attributes(tf)
```

### ...from a `move2` object

```{r}
library(move2)

# Load a move2 object using the move2 package
albatross_move2 <- mt_read(mt_example())
albatross_move2 <- albatross_move2[
  !sf::st_is_empty(albatross_move2),
] |> sf::st_transform(suggest_utm_crs(albatross_move2))

# Convert to trackframe
albatross_tf <- as.trackframe(data = albatross_move2)
class(albatross_tf)
```

### ...from an `sftrack` object

```{r}
library(sftrack)

# Load and prep an sftrack object
data("raccoon", package = "sftrack")
raccoon$month <- as.POSIXlt(raccoon$timestamp)$mon + 1
raccoon$time <- as.POSIXct(raccoon$timestamp, tz = "EST")
coords <- c("longitude", "latitude")
group <- list(id = raccoon$animal_id, month = as.POSIXlt(raccoon$timestamp)$mon + 1)
time <- "time"
error <- "fix"
crs <- 4326

my_sftrack <- as_sftrack(
  data = raccoon,
  coords = coords,
  group = group,
  time = time,
  error = error,
  crs = crs
) |> sf::st_transform(suggest_utm_crs(raccoon$latitude, raccoon$longitude))

# Convert to trackframe
sftrack_tf <- as.trackframe(my_sftrack)
class(sftrack_tf)
attributes(sftrack_tf)
```

## Extract specific columns from a `trackframe`

**trackframe** has specific fuctions to extract the most important columns from the a `trackframe` object, regardless of the actual names of those columns. The following four functions extract the easting column (i.e., x or longitude), the northing column (i.e., y or latitude), the time column, and the animal id column, respectively.

```{r}
easting(tf)
```

```{r}
northing(tf)
```

```{r}
time(tf)
```

```{r}
id(tf)
```

There is a also a function to extract a vector of the unique values in the animal id column:

```{r}
unique_ids(tf)
```

## Subset and divide a `trackframe` by `id`

To extract the track of a specific animal, use the `select_id` function.

```{r}
tf_2 <- select_id(tf, "2")
tf_2
```

To split the tracks of each individual, and store them as a list of `trackframe` objects, use the `split_by_id` function.

```{r}
tf_split <- split_by_id(tf)
class(tf_split)
names(tf_split)
tf_split[[1]]
tf_split[["2"]]
```

## Extract specific column names from a `trackframe`

FIXME: Insert explainer for [#89](https://github.com/livingingroups/travelpaths-devel/issues/89) and [#91](https://github.com/livingingroups/travelpaths-devel/issues/91), once they've been implemented.

## Transform from a `trackframe`

To transform a `trackframe` object into another type of animal track data, there are two approaches.

The first approach takes any `trackframe` object and transforms it into a specific other animal track data type using the `tf_as_*` functions. In this case, the output object will have the same CRS as the `trackframe` object.

The second approaches takes a `trackframe` object *that was created from another type of animal track data using the* `as.trackframe` *function*, and transforms it back into its original object type using the `tf_backtransform` function. In this case, the output object will have the same CRS as the original input object.

The first approach can be used on any `trackframe` object, regardless of its history. The second approach can only be used on `trackframe` objects that were created using the `as.trackframe` function.

### ... to a `data.frame`

FIXME: Logic from [#87](https://github.com/livingingroups/travelpaths-devel/issues/87) related to backtransformation should go here once that logic is implemented.

### ... to a `move2` object

```{r}
# First approach (ignoring that this object was originally a move2 obj)
ap1_recovered_move2 <- tf_as_move2(albatross_tf)

# Second approach (backtransforming this object to its original type, i.e., a move2 obj)
ap2_recovered_move2 <- tf_backtransform(albatross_tf)

head(ap1_recovered_move2) # <- points have same CRS as albatross_tf
head(ap2_recovered_move2) # <- points have same CRS as original albatross_move2
```

### ... to an `sftrack` object

```{r}
# First approach
ap1_recovered_sf <- tf_as_sftrack(sftrack_tf)

# Second approach
ap2_recovered_sf <- tf_backtransform(sftrack_tf)

head(ap1_recovered_sf)
head(ap2_recovered_sf)
head(my_sftrack)
```

## Example Workflow

The following example shows how to create a function that will add a "step length" column to a `trackframe` object, and then create a `trackframe` object from a move2 object, split it by animal id, add the "step length" column, then transform the object back into a move2 object.

It is advantageous to use the **trackframe** package for this, as it allows the `step_length()` function to be easily applied to any animal movement data structure, by simply converting the data to `trackframe`, rather than having to alter the function itself.

First, create the `step_length()` function:

```{r}
step_length <- function(x, y)  {
  sqrt(diff(x)^2 + diff(y)^2)
}
```

Because this function operates directly on the location coordinates themselves, it's impossible to apply it directly to a move2 object (which stores the location coordinates in a special geometry column). So, we can easily convert the move2 object to a trackframe object:

```{r}
tf <- as.trackframe(albatross_move2)

# sort by id then by time (can use regular dataframe operators)
tf <- tf[order(id(tf), time(tf)), ]
```

As a trackframe object, we now have direct access to the coordinates, and so we can apply the `step_length()` function:

```{r}
# now, extract all of the step lengths
step_length_vec <- step_length(easting(tf), northing(tf))
# (Length of vector is one less than the number of rows in the tf, because the
# first location doesn't have a step length)

# need to also replace the first "step" of each individual with NA (since this is
# not a real step, just the distance between one indivs last location and another
# indivs first location)
step_length_vec[which(id(tf)[1:(nrow(tf) - 1)] != id(tf)[2:nrow(tf)])] <- NA

# now attach the new vector to the tf (and put NA in first row)
tf$step_length <- c(NA, step_length)

# take a look at it ######UPDATE ONCE THIS EXISTS
##tf[1345:1355, c(time(tf), id(tf), "Step_length")]
cbind(time(tf), id(tf), tf[, "step_length"])[1345:1355, ]#temp fix
```

\^\^ FIXME: Update 'take a look code', once [#89](https://github.com/livingingroups/travelpaths-devel/issues/89) and [#91](https://github.com/livingingroups/travelpaths-devel/issues/91), have been implemented.

Then we can convert the whole thing back to a move2 object if needed:

```{r}
albatross_move2_w_step_length <- tf_backtransform(tf)
```

Using **trackframe** would also allow you to wrap the entire thing into a function, which could then be applied to animal movement data of any structure (e.g., move2, sf_track, etc), for example:

```{r}
add_step_length_col <- function(x) {
  tf <- as.trackframe(x)
  tf <- tf[order(id(tf), time(tf)), ]
  step_length_col <- step_length(easting(tf), northing(tf))
  step_length_col[which(id(tf)[1:(nrow(tf) - 1)] != id(tf)[2:nrow(tf)])] <- NA
  tf$step_length <- c(NA, step_length_col)
  tf_backtransform(tf)
}
```
