---
author: "Trackframe Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: Quickstart Guide
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Quickstart Guide}
---

# Quickstart Guide

This vignette demonstrates the main functions of trackframe and how to get started using this package. For a full description of the benefits of using trackframe over other approaches, see `vignette("why_trackframe")`

## Load Package

```{r}
library(trackframe)
```

## Create a `trackframe`

It's quite simple to convert common animal track data representations into `trackframe` objects. When creating a `trackframe` object from another data representation, `as.trackframe` identifies the columns representing time (`time_col`), coordinates (`northing_col` & `easting_col`), and animal id (`id_col`). It obtains this information from 3 potential sources: 1) metadata inherited from the input object, 2) arguments supplied directly to `as.trackframe`, and/or 3) pre-set global package options. For details about the process and specifications of identifying the key columns, see `vignette("identifying_columns")`.

`as.trackframe` also sorts the data that it is given, firsb by the values in the `id_col` and then by the values in the `time_col`. To prevent this default sorting, you can set `sort = FALSE` in the `as.trackframe` function.

### ...from a non-sf `data.frame`

Convert a plain (base R) `data.frame` to a `trackframe` object, first by specifying the column names in the arguments of `as.trackframe` and then via the global options:
```{r}
# Create example data
df <- data.frame(
  timestamp = rep(as.POSIXct(Sys.time() + 1:5), times = 5),
  x = runif(25, 0, 10),
  y = runif(25, 0, 10),
  id = rep(5:1, each = 5)
)

# Convert to trackframe using column names supplied to the relevant arguments
# to identify key columns
tf_args <- as.trackframe(
  df,
  time_col = "timestamp",
  easting_col = "x",
  northing_col = "y",
  id_col = "id",
  crs = "EPSG:32632"
)

# Convert to trackframe using the pre-set global package options to identify
# the key columns
tf_go <- as.trackframe(df,
  crs = "EPSG:32632"
)

#in this case, tf_args is the same as tf_go
identical(tf_args, tf_go)
head(tf_args)
head(tf_go)
attributes(tf_args)[1:6]
attributes(tf_go)[1:6]
```

Now take a look at how `as.trackframe` sorting works:
```{r}
# As above, convert to trackframe using column names supplied to the relevant arguments
# to identify key columns, with sort = TRUE (the default)
tf_args_sorted <- as.trackframe(
  df,
  time_col = "timestamp",
  easting_col = "x",
  northing_col = "y",
  id_col = "id",
  crs = "EPSG:32632",
  sort = TRUE
)

tf_args_unsorted <- as.trackframe(
  df,
  time_col = "timestamp",
  easting_col = "x",
  northing_col = "y",
  id_col = "id",
  crs = "EPSG:32632",
  sort = FALSE
)

# Compare sorted vs unsorted
tf_args_sorted
tf_args_unsorted
```

### ...from a `move2` object

When applying `as.trackframe` to a move2 object, key columns are identified using the metadata of the input object. Because `as.trackframe` can only handle Cartesian (i.e., projected) coordinates, objects that are in non-projected coordinates (e.g., latitude and longitude) must first be transformed. For more details about compatible coordinate reference systems, see `vignette("crs")`.

```{r}
library(move2)

# Load a move2 object using the move2 package, and project the coordinates to UTMs
albatross_move2 <- mt_read(mt_example())
albatross_move2_proj <- albatross_move2[
  !sf::st_is_empty(albatross_move2),
] |> sf::st_transform(crs = 3857)

# Convert to trackframe, key columns and crs will be identified using metadata of the
# input move2 object
albatross_tf <- as.trackframe(data = albatross_move2_proj)

# Check out the new trackframe object
class(albatross_tf)
attributes(albatross_tf)[1:6]
str(albatross_tf)
```

<!-- You can use the `suggest_utm_crs` function to identify the EPSG code for the most likely UTM zone to which you should project your data.

[ONCE THIS IS IMPLEMENTED, remove code block above, and activate code block below.]

```{r, eval=FALSE, include = FALSE}
library(move2)

# Load a move2 object using the move2 package, and project the coordinates to UTMs
albatross_move2 <- mt_read(mt_example())
albatross_move2_proj <- albatross_move2[
  !sf::st_is_empty(albatross_move2),
] |> sf::st_transform(suggest_utm_crs(albatross_move2))

# Convert to trackframe, key columns and crs will be identified using metadata of the
# input move2 object
albatross_tf <- as.trackframe(data = albatross_move2_proj)

# Check out the new trackframe object
class(albatross_tf)
crs(albatross_tf)
attributes(albatross_tf)
str(albatross_tf)
```

-->

### ...from an `sftrack` object

As with `move2` objects, when applying `as.trackframe` to an `sftrack` object, key columns are identified using the metadata of the input object. Because `as.trackframe` can only handle Cartesian (i.e., projected) coordinates, objects that are in non-projected coordinates (e.g., latitude and longitude) must first be transformed. For more details about compatible coordinate reference systems, see `vignette("crs")`.

```{r}
library(sftrack)

# Load and prep an sftrack object
data("raccoon", package = "sftrack")
raccoon$month <- as.POSIXlt(raccoon$timestamp)$mon + 1
raccoon$time <- as.POSIXct(raccoon$timestamp, tz = "EST")
coords <- c("longitude", "latitude")
group <- list(id = raccoon$animal_id, month = as.POSIXlt(raccoon$timestamp)$mon + 1)
time <- "time"
error <- "fix"
crs <- 4326

my_sftrack <- as_sftrack(
  data = raccoon,
  coords = coords,
  group = group,
  time = time,
  error = error,
  crs = crs
) |> sf::st_transform(crs = 3857)

# Convert to trackframe, key columns and crs will be identified using metadata of the
# input sftrack object
sftrack_tf <- as.trackframe(my_sftrack)
class(sftrack_tf)
attributes(sftrack_tf)[1:6]
```

<!-- You can use the `suggest_utm_crs` function to identify the EPSG code for the most likely UTM zone to which you should project your data. 

[ONCE THIS IS IMPLEMENTED, remove code block above, and activate code block below.]

```{r, eval=FALSE, include = FALSE}
library(sftrack)

# Load and prep an sftrack object
data("raccoon", package = "sftrack")
raccoon$month <- as.POSIXlt(raccoon$timestamp)$mon + 1
raccoon$time <- as.POSIXct(raccoon$timestamp, tz = "EST")
coords <- c("longitude", "latitude")
group <- list(id = raccoon$animal_id, month = as.POSIXlt(raccoon$timestamp)$mon + 1)
time <- "time"
error <- "fix"
crs <- 4326

my_sftrack <- as_sftrack(
  data = raccoon,
  coords = coords,
  group = group,
  time = time,
  error = error,
  crs = crs
) |> sf::st_transform(suggest_utm_crs(raccoon$latitude, raccoon$longitude))

# Convert to trackframe, key columns and crs will be identified using metadata of the
# input sftrack object
sftrack_tf <- as.trackframe(my_sftrack)
class(sftrack_tf)
attributes(sftrack_tf)
```
-->


<!-- ### ...from an `sf` object

ISSUE #139 

Add an example of a regular sf object... at least an sf POINT dataframe, maybe also MULTIPOINT or MULTILINESTRING sf dataframes. -->


## Extract and/or replace key columns from a `trackframe`

Trackframe has specific functions to extract or replace values in the key columns from the a `trackframe` object, regardless of the actual names of those columns. The following four functions get the values in the `time_col`, `easting_col`, `northing_col` and `id_col`, and can be used to extract and/or replace values in those columns.

```{r}
# Create example data
df <- data.frame(
  timestamp = rep(as.POSIXct(Sys.time() + 1:5), times = 5),
  x = runif(25, 0, 10),
  y = runif(25, 0, 10),
  id = rep(1:5, each = 5)
)

# Convert to trackframe
tf <- as.trackframe(df,
  crs = "EPSG:32632")
```


```{r, include=FALSE}
# Extract the values in the time_col
time(tf)

# Replace the second value in the time_col
time(tf)[2] <- as.POSIXct("2025-11-17 14:08:00 CET")

tf
```

```{r, include=FALSE}
# Extract the values in the easting_col
easting(tf)

# Replace all values in the easting_col with a new vector
easting(tf) <- runif(25, 0, 10)

tf
```

```{r, include=FALSE}
# Extract the values in the northing_col
northing(tf)

# Replace every values in the northing_col with a 5
northing(tf) <- 5

tf
```

```{r, include=FALSE}
# Extract the values in the id_col
id(tf)

# Replace the values in the id_col based on a conditional
id(tf) <- ifelse(id(tf) == 1, "Bob", "Other")

tf
```
-->

There is a also a function to extract a vector of the unique values in the animal id column:

```{r}
unique_ids(tf)
```

## Subset and divide a `trackframe` by `id`

To extract the track of a specific animal, use the `select_id` function.

```{r}
tf_2 <- select_id(tf, "Bob")
tf_2
```

To split the tracks of each individual, and store them as a list of `trackframe` objects, use the `split_by_id` function.

```{r}
tf_split <- split_by_id(tf)
class(tf_split)
names(tf_split)
tf_split[[1]]
tf_split[["Bob"]]
```

- *_col (e.g. "easting_col") gets / sets the key column's column name. In the "set" case,
  it updates the designated column's name. It does not change which column is designated as a key column.
- "easting"/"northing"/etc. gets / sets the corresponding values in the dataframe
- `as.trackframe` can be re-run to change which columns are treated as key columns


## Transform from a `trackframe`

To transform a `trackframe` object into another representation of animal track data, there are two approaches.

The first approach takes any `trackframe` object and transforms it into another animal track data representation using the `tf_as_*` functions. In this case, the output object will be sorted in the same order as the `trackframe` object.

The second approach takes a `trackframe` object *that was created from another representation of animal track data using the* `as.trackframe` *function*, and transforms it *back* into its original data representation using the `tf_backtransform` function. In this case, if sorting was performed during the conversion of the original object in `as.trackframe` will be undone.

The first approach can be used on any `trackframe` object, regardless of its history. The second approach can only be used on `trackframe` objects that were created using the `as.trackframe` function.

### ... to a non-sf `data.frame`

```{r}
# First approach (ignoring that this object was originally a df obj)
df1_recovered <- tf_as_xyt(tf_args)
head(df1_recovered)
attributes(df1_recovered)

# Second approach (backtransforming this object to its original type, i.e., a dataframe obj)
df2_recovered <- tf_backtransform(tf_args)
head(df2_recovered)
attributes(df2_recovered)

# Compare to original df
head(df)
attributes(df)
```

Note that `df1_recovered` (the dataframe created by `tf_as_xyt`) is sorted in ascending order by `id` and `timestamp` (this sorting that was performed when it was first converted to a `trackframe` object, and has now been maintained), whereas `df2_recovered` (the dataframe created by `tf_backtransform`) is in the order of the original, pre-conversion-to-trackframe, `dataframe` object.  

### ... to a `move2` object

```{r}
# First approach (ignoring that this object was originally a move2 obj)
ap1_recovered_move2 <- tf_as_move2(albatross_tf)

# Second approach (backtransforming this object to its original representation, i.e., a move2 obj)
ap2_recovered_move2 <- tf_backtransform(albatross_tf)

# Compare with each other and original object
head(ap1_recovered_move2)
head(ap2_recovered_move2)
head(albatross_move2_proj)
```

### ... to an `sftrack` object

```{r}
# First approach
ap1_recovered_sf <- tf_as_sftrack(sftrack_tf)

# Second approach
ap2_recovered_sf <- tf_backtransform(sftrack_tf)

# Compare with each other and original object
head(ap1_recovered_sf)
head(ap2_recovered_sf)
head(my_sftrack)
```

Here again you can see evidence of `as.trackframe`'s default `sort = TRUE`. Sorted: `ap1_recovered_sf` (the dataframe created by `tf_as_sftrack`) is sorted in ascending order by `id` and `timestamp`, with animal id `TTP-041` appearing in the `head()` call. This is the order that the data were sorted into when it was first converted to an `as.trackframe` object. Unsorted: `ap2_recovered_sf` (the dataframe created by `tf_backtransform`) has been "unsorted" -- animal id `TTP-058` appears first in the `head()`. This is the order of the original, pre-conversion-to-trackframe, `sftrack` object.


<!-- ### ...to an `sf` object

ISSUE #139 

Add an example of a regular sf object... at least an sf POINT dataframe, maybe also MULTIPOINT or MULTILINESTRING sf dataframes. -->

## Example Workflow

The following example shows how to use `trackframe` to create a function that can take any representation of animal track data as input, and will add a "step length" column to the data.

Within the function, you can convert the data to a `trackframe` object, then apply the desired calculations, then convert the data back to its original representation. The trackframe approach requires far fewer, and simpler, lines of code than writing the function in such a way that it can recognize and accommodate each of multiple different data representations. Furthermore, because `trackframe` stores coordinate columns as plain numeric vectors, it's easy to write functions that employ basic geometry and math to perform calculations directly on the location coordinates values themselves. See `vignette("why_trackframe")`. for more information on the design of the `trackframe` object.

```{r}
# Function to add a step length column to a animal track data object
add_step_length_col <- function(x) {
  # convert input data object to a tf object
  tf <- as.trackframe(x)
  # sort by animal id then time
  tf <- tf[order(id(tf), time(tf)), ]
  # use basic geometry to calculate the distance between successive locations
  step_length_col <- sqrt(diff(easting(tf))^2 + diff(northing(tf))^2)
  # replace the first "step" of each individual with NA
  step_length_col[which(id(tf)[1:(nrow(tf) - 1)] != id(tf)[2:nrow(tf)])] <- NA
  # attach step length column to tf (with preceeding NA for first location)
  tf$step_length <- c(NA, step_length_col)
  # transform the tf back into its orginal input representation
  tf_backtransform(tf)
}

# apply the function to a projected move2 object
albatross_move2_proj2 <- add_step_length_col(albatross_move2_proj)

# take a look at it
head(albatross_move2_proj2)
tail(albatross_move2_proj)

# look at rows where the animal id changes
albatross_move2_proj2[1345:1355, c("timestamp", "individual-local-identifier", "step_length")]
```

Without changing the function at all, we can apply `add_step_length_col` to another animal track data representation, such as an `sftrack` object.

```{r}
# apply the function to a projected sftrack object
my_sftrack <- add_step_length_col(my_sftrack)

# take a look at it
head(my_sftrack)
tail(my_sftrack)
```
