---
author: "TODO"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples of Trackframe}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  chunk_output_type: console
---

# Load Package

```{r}
library(trackframe)
```

# Create a `track_frame`

## General

In an object of class `track_frame` we always assume to have UTM easting/northing as coordinates.
`as.track_frame` automatically transform coordinates to UTM easting/northing.

## from a `data.frame`

```{r}
df <- data.frame(
  time_col = as.POSIXct(Sys.time() + 1:5),
  easting_col = runif(5, 0, 10),
  northing_col = runif(5, 0, 10),
  id = 1:5
)
tf <- as.track_frame(df, time_col = "time_col", easting_col = "easting_col",
                        northing_col = "northing_col", id_col = "id")
head(tf)
attributes(tf)
```

## from a `move2` object

```{r}
library(move2)
albatross_move2 <- mt_read(mt_example()) |>
  sf::st_transform(3857)
albatross_move2 <- albatross_move2[!sf::st_is_empty(albatross_move2),]
albatross_tf <- as.track_frame(albatross_move2)
class(albatross_tf)
```

Backtransform `track_frame` to `move2` object

```{r}
recovered_move2 <- tf_as_move2(albatross_tf[!is.na(northing(albatross_tf)),])
str(recovered_move2)
```

## from an `sftrack` object

```{r}
library(sftrack)
library(trackframe)
data("raccoon", package = "sftrack")
raccoon$month <- as.POSIXlt(raccoon$timestamp)$mon + 1
raccoon$time <- as.POSIXct(raccoon$timestamp, tz = "EST")
coords <- c("longitude","latitude")
group <- list(id = raccoon$animal_id, month = as.POSIXlt(raccoon$timestamp)$mon+1)
time <- "time"
error <- "fix"
crs <- 4326
# create a sftrack object
my_sftrack <- as_sftrack(data = raccoon,
                         coords = coords,
                         group = group,
                         time = time,
                         error = error,
                         crs = crs)

sftrack_tf <- as.track_frame(my_sftrack)
class(sftrack_tf)
attributes(sftrack_tf)
```

Backtransform `track_frame` to `sf` object

```{r}
recovered_sf <- tf_as_sftrack(sftrack_tf[!is.na(northing(sftrack_tf)),])
str(recovered_sf)
```


## from `cocomo` style data

```{r}
library(travelpaths)
set.seed(2025)
cocomo <- tf_as_cocomo(sim_travel_paths(3, 3))
str(cocomo)
cocomo$x
cocomo$y
cocomo$t
cocomo$ids

cocomo_tf <- cocomo_as_tf(cocomo$x, cocomo$y, cocomo$t, cocomo$ids)
cocomo_tf
class(cocomo_tf)


```

Retransform `track_frame` to cocomo style data

```{r}
cocomo2 <- tf_as_cocomo(cocomo_tf)
str(cocomo2)
```



# Extract information

```{r}
tf <- sim_travel_paths(3, 5)
```


```{r}
easting(tf)
```

```{r}
northing(tf)
```

```{r}
time(tf)
```

```{r}
id(tf)
```



# Simulate paths

Generate a single track in `matrix` format

```{r}
data_matrix <- sim_travel_path(100, format = "matrix")
```

Generate a single track in a `data.frame` format

```{r}
data_df <- sim_travel_path(100, format = "data.frame")
```

Generate a single track in a `track_frame` format

```{r}
tf <- sim_travel_path(100, format = "track_frame")
```

Generate 3 tracks with different sizes

```{r}
ntracks <- 3
sizes <- c(2, 4, 5)
multi_track <- sim_travel_paths(ntracks, sizes)
```

Generate 5 tracks all with the same size

```{r}
uniform_tracks <- sim_travel_paths(5, 10)
```

Extract a specific track

```{r}
track2 <- select_id(multi_track, "track_2")
```

